Pool Model - A unified Mathematical\\ Modeling Framework for Food Science

### 2025-11-06 paper/pool_model.tex
delete commented old text

% \paragraph{General Introduction to Bacterial Consortia}
% \begin{itemize}
%     \item Complex system; many players (cell-types, interactions)
%     \item processes on different scales (population, cellular, molecular, gene-level)
%     \item Food Safety, Production
%     \item Healthcare: Drug development
%     \item Environmental Sustainabiliy
% \end{itemize}
% Motivation: general, flexible approach for bacterial systems that allows easily modify the system depending on processes on different scales.
% Allows us to describe the processes within populations using different bacterial states as well as interactions between different populations using ODEs.

Bacterial consortia are highly complex systems which consist of multiple distinct cell types.
The underlying processes are interconnected with each other and occur on all scales, ranging from
population-based dynamics of colony growth over molecular processes such as metabolism to gene
expression.
Bacterial communities are ubiquitous and even able to survive under extreme conditions, making them
virtually omnipresent around any living system.
They play an especially important role within the context of food safety and production
being a key contributor to the spoilage of food products as well as fermentation processes.
Furthermore, effects such as biofilm formation can hinder drug treatment by shielding the bacterial
colony from antibiotic substances or immune response of the organism.
The natural bacterial processes are a crucial part of our planet's ecosystem, being responsible for
tasks such as nutrient cycling, decomposition of organic matter and oxygen production.
It is of great interest to understand the dynamics of these various processes.

% \paragraph{Experimental/Biological evidence/phenomena}
% \begin{itemize}
%     \item Lag Phase, Competition, Cooperation, Limiting Resource
%     \item Comprehensive reviews about competition/cooperation bacterial mechanisms: \citep{ghoul_ecology_2016, stubbendieck_bacterial_2016, hibbing_bacterial_2010, west_social_2007}
%     \item Spatial Effects: adhesion, migration, chemotaxis, diffusion
%     \item Quorum-sensing
%     \item Biofilm formation \citep{Jin2020}
% \end{itemize}

In order to achieve a comprehensive description of the target system, it is important to incorporate
all relevant biological phenomena arising in these bacterial systems.
During their reproductive process, bacteria undergo multiple phases~\citep{buchanan_when_1997}, starting in a lag phase which
is necessary to adapt to new environments and initially impedes the overall growth~\citep{rolfe_lag_2012, monod_growth_1949}.
Additionally, bacteria are able to cooperate, e.g., via complementary metabolic processes or inhibit
each other, e.g., by means of waste production as a metabolic byproduct~\citep{west_social_2007, rainey_evolution_2003, hibbing_bacterial_2010, stubbendieck_bacterial_2016}.
A limited supply of nutrients will also affect the overall reproductive capacity of the colony~\citep{monod_growth_1949}.
Bacteria have thus devised various survival strategies to ensure sufficient access to nutrients.
Effects on a single-cell level such as adhesion~\citep{htuson_bacteriasurface_2013}, migration~\cite{decoene_microscopic_2011} and chemotaxis can lead to complex
phenomena on a population-based level which can aid in the goal of survival.
Cell signaling, e.g., in the form of quorum-sensing can lead to collective emergent phenomena such as
biofilm formation which play an important role for many mature microbial communities~\citep{Jin2020,ng_bacterial_2009}.

% \paragraph{ODE models (predictive microbiology, existing, overview, most popular)}
% Existing models in Predictive Microbiology (Primary Models)~\citep{perez-rodriguez_predictive_2012}:
% \begin{itemize}
%     \item An overview of primary models used in food science can be found in~\citep{van_boekel_kinetic_2008}.
%     \item Baranyi-Roberts Model~\citep{baranyi_modeling_1993, baranyi_dynamic_1994}
%     \item modified Gompertz~\citep{Gompertz1825-wi,Gibson1987}
%     \item Buchanan et al.: three-phase linear model for lag phase, exponential growth phase, and stationary phase~\citep{buchanan_when_1997}
%     \item McKellar two-compartment model with bacteria in growing and non-growing state~\citep{mckellar_heterogeneous_1997}
%     \item Inactivation models
% \end{itemize}
% %
% Secondary models:
% \begin{itemize}
%     \item Polynomial models 
%     \item Square Root-Type Models, e.g., Ratkowski for temperature~\citep{ratkowsky_relationship_1982}
%     \item The Cardinal Parameter Model~\citep{zwietering_decision_1992}
% \end{itemize}

The first step in the construction of any mathematical model is to determine the required level of
detail with respect to the description scale and consequently which effects to include.
The bacterial growth curves let us describe the dynamics of the overall abundance of various
species.
The field of predictive microbiology has developed a variety of ODE-based models to capture
bacterial growth dynamics~\citep{perez-rodriguez_predictive_2012}.
Various \acp{ode} \textit{primary models} describe the bacterial growth curve under constant
environmental conditions.
Widely used examples include the modified Gompertz model~\citep{Gompertz1825-wi,Gibson1987}, Baranyi-Roberts
model~\citep{baranyi_modeling_1993, baranyi_dynamic_1994} and McKellar two-compartment model~\cite{buchanan_when_1997} which distinguishes between growing and non-growing bacteria.
To take into account changing environmental conditions, so-called \textit{secondary models}
formulate the model parameters as a function of external quantities such as temperature, pH or
humidity.
These can be simple polynomial models, square root-type models~\citep{ratkowsky_relationship_1982} or
more complex relations such as the cardinal parameter model~\citep{zwietering_decision_1992}.
The advantage of the \ac{ode} formulation lies in the predictive power and simplicity of usage.
However, due to their coarse-grained nature, they lack the ability to describe heterogeneity and
other spatial effects.

% \paragraph{Models for spatial effects; PDEs, \acp{abm}, others}
% \begin{itemize}
%     \item BSim \citep{Gorochowski2012,Dang2020}
%     \item Bacterial Branching patterns \citep{Kawasaki1997,Matsushita1998}
% \end{itemize}

% To incorporate spatial heterogeneity or other effects other modeling approaches have been devised.
% \acp{pde} follow a similar pattern as \acp{ode} but also distribute the described variables through
% space.
When dealing with spatial effects such as heterogeneity, we are required to use more extensive
modeling formulations such as \acp{pde} or \acp{abm}.
The former represent an intermediate step between the population-based \ac{ode} models and the
individually-based \acp{abm} and can be effective, especially when dealing with large crowds of
cells that can be approximated as a continuum.
On the other hand, \acp{abm} excel at describing the individual behavior of cells which can then
lead to more complex collective phenomena.
Tools such as BSim~\citep{Gorochowski2012,Dang2020}, AgentCell~\cite{emonet_agentcell_2005}, BacSim~\cite{kreft_bacsim_1998} or \texttt{cellular\_raza}~\cite{Pleyer2025} are
capable to describe large amounts of bacteria, performing various intracellular and extracellular
processes.
The latter has been used to adapt \ac{pde}-based models into an \ac{abm} formulation which describe
bacterial branching patterns~\citep{Kawasaki1997,Matsushita1998}.

% \paragraph{What is the problem/what is the question? How do we address this?}

In this work, we introduce a new \ac{ode}-based modeling framework which describes different
bacterial species, their metabolic states and external quantities such as nutrients as individual
pools.
The dynamics of bacterial growth can be understood in terms of transitions between the different
bacterial states, i.e., pools.
Our model also allows us to capture interactions between these bacterial states and the external
microenvironment.
Compared against the aforementioned methods, this approach turns out to be less complex with general
applicability, low number of parameters and greater explainability for the underlying biological
reality.
To illustrate the generality of the model, multiple examples are reported.
We analyze the spatial limitations of an interacting two-species model and show that it can be fully
understood on an individual-based level in terms of single-cell processes.
Furthermore, we assess how spatial heterogeneity~\citep{mckellar_heterogeneous_1997} can impact the shape of the growth curves.

@@

% TODO change the name "death pool"?

@@

% The dynamics of the three pool model can be captured by the coupled system of ordinary differential

@@

% That means that as long as there are bacteria in the growth phase $G_B>0$, toxin is produced, no matter whether the bacterial count $N_B$ is still growing.

@@

\section{Spatial effects}
As described above, the proposed model operates with the total number of the bacterial counts for
different species and states in the system, not accounting for their spatial distributions.
Assuming that the bacterial system is uniformly distributed (sometimes also called 'well-stirred')
we expect the \ac{ode} description to be valid.
However, in reality spatial effects can play a significant role for population survival forcing some
species to develop competitive mechanisms such as cell adhesion to surfaces in the areas rich with
nutrients~\citep{htuson_bacteriasurface_2013}.
Another mechanism is the motility of bacteria, e.g., the flagellates of \textit{E. coli} or
\textit{B. subtilis} which are capable of chemotactic motion driven by oxygen
concentrations~\citep{decoene_microscopic_2011}.
The bacterial social interactions such as release of toxins or public goods can also
significantly affect the spatial structure of the community~\citep{blanchard_bacterial_2015}.
These effects can only be treated within a spatial framework that is able to describe the observed
effects.\\
Compared to other modeling techniques such as \acp{pde}, \acp{abm}~\citep{nagarajan_agent-based_2022-1} force researchers to think in
terms of cellular processes, thus bridging the gap from a
collective (bulk) model to an individual-based (single-cell) description.
Our goal is to provide an \ac{abm} which can be used to derive the given \ac{ode} model as a
coarse-grained case.
In this way, we can show that the bulk-based \ac{ode} model is directly related to the mechanistic
single-cell behavior of the bacteria.
To this end, we consider the system consisting of two bacterial species that can be described by the two internal states: lag phase and growth.
Both of these species grow using the same resource pool $R$ but with different growth constants
$\alpha_{A/B}$ and different lag phase transition rates $\lambda_{A/B}$.
Moreover, species $A$ is able to release a certain antimicrobial component $I$ that shows an
inhibitory effect on species $B$ (amensalism).
We assume that $L,G$ are measured in terms of the overall volume and $R,I$ are measured as total
amount.
The \acp{ode} system of the bacterial population can be written as:
% - \citep{emonet_agentcell_2005, gorochowski_bsim_2012, li_nufeb_2019, kreft_bacsim_1998}: examples of another toolboxes for bacterial communities modeling (ABM).\\
% More about the modeling of the motile bacteria can be seen in works:~\citep{decoene_microscopic_2011, rosser_modelling_2014, sokolov_physical_2012, li_amplified_2008}.
% In general, bacterial communities in 2D can experience collective motion called swarming~\citep{wu_collective_2015}.
% Some works modelled 2D motile bacteria:
% \begin{itemize}
%     \item \citep{decoene_microscopic_2011}: 2D chemotactic self-propelled bacterial motion of the flagellated bacteria (e.g., E. coli, B. subtilis).
%     Chemotaxis here was driven by the oxygen concentrations.
%     \item \citep{rosser_modelling_2014}: modeling of the bacterial reorintation, rotation.
%     \item \citep{sokolov_physical_2012, li_amplified_2008}: modeling of the bacteria using hydrodynamic interactions and Brownian motion.
%     Fluctuations in velocity and rotations.
%     (\citep{li_amplified_2008}: Brownian motion gives right motion observed using microscopy).
%     \item \citep{farrell_mechanically_2013}: 2D modeling using Newtonian dynamics and only mechanical interactions + nutrient diffusion.
%     Branching parameter defines diff limits where diffusion dominant and limits growth, and where it's irrelevant.
%     Mechanical pressure (due to bacteria-bacteria interactions) affects growth, can limit it.(!!!)
% \end{itemize}
% For simulation results:
% \begin{itemize}
%     \item \citep{farrell_mechanically_2013}: Showed the influence of mechanic interactions between bacteria in 2D system with bacteria consuming the nutrient.
%     The growth of bacterial colonies where non-motile microorganisms replicate and push each other away as they grow.
%     Found a transition between two different growth regimes, controlled by the balance between growth and uptake of nutrient:\\
%     Exponential growth where diffusion of the nutrient does not play any role. Sublinear growth where growth is limited by nutrient diffusion.
%     (Structurally in 2D it's regimes: smooth propagating of bacterial front vs. branching.)
%     \item \citep{nagarajan_agent-based_2022-1}: review on the ABM for bacterial population.
%     The ABM represent a microbial colony as a group of discrete agents that are governed by a set of rules, where each agent can be an individual cell or a collection of cells.
%     Can use citation when talk about general processes we are accounting as growth, uptake of nutrient, production of inhibitor, e.g.,:
%     Physical interactions between agents are calculated using excluded volume interactions.
%     Chemical interactions as releasing inhibitors are described using concentrations and solved using PDEs.
% \end{itemize}
        \dot{G}_A &= \lambda_A R L_A + \alpha_A R G_A\\
        \dot{G}_B &= \lambda_B R L_B + \frac{\alpha_B}{1 + \mu_I I} R G_B.
    \label{eq:spatial_limit_F}
Moreover, the microenvironment of our system contains the resource and the inhibitor molecules
        \dot{R} &= -\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{(1 + \mu_I I) N_t} R G_B \\
        \dot{I} &= \kappa G_A.
    \label{eq:spatial_limit_H}

% ------------------------------------------------------------------------------------------------
\subsection{Constructing an agent-based pool model}
The following sections will demonstrate how a given \ac{ode} system can be translated into a
custom-built \ac{abm} which has only a few more overall parameters.
This procedure is a blueprint which can be followed to construct \acp{abm} for similar
\ac{ode}-based systems.\\
To construct our \ac{abm}, we use the modeling framework \texttt{cellular\_raza}~\citep{Pleyer2025} that allows us to define our bacterial agents from a clean slate.
With this flexibility, researchers are able to construct minimal models using only as many parameters which are necessary.
This ability, to precisely define and control the exact behavior of our cell-agents, enables the following parts.

@@

% How well does the pool model do to encapsulate all the different growth models
% \begin{itemize}
%     \item All 6 models can be represented by pool model
%     \item Baranyi-Roberts results can be reproduced, but not identical; requires some work
%     \item what are some general limitations within this framework? how can they be circumvented?
%     \item why is the pool model "better" than the other \ac{ode} models? (why did we do this?);
%         generalization of many models; better explainability;
%     \item link to supplement; we can generalize this and prove theorems for the abstract,
%         generalized model; this allows us to discuss many cases at once
% \end{itemize}
%
% How can we cover more effects of bacteria?
% \begin{itemize}
%     \item More states for metabolic processes
%     \item More states for more species
%     \item Other cooperiation/inhibition effects
%     \item external influences
% \end{itemize}
%
% How justified was the construction of the \ac{abm}?
% \begin{itemize}
%     \item Discuss simplified physical interaction model
%     \item Heterogeneity plays a role; as we have shown
%     \item why did we not simply use a PDE?; because we have to think in terms of cellular processes
%         and this ensures that our model is actually realistic
% \end{itemize}
%
% What are limitations of our approach?
% \begin{itemize}
%     \item spatial stuff; heterogeneity (because \ac{ode})
%     \item larger models require more parameters, harder to estimate
%     \item only simple models have analytical solutions
% \end{itemize}

The presented pool model was used to describe the dynamics of one- and two-species systems containing the main growth stages and interactions between bacteria in the form of cooperation and competition.
Our approach produces results which exhibit similarities to other well-known predictive microbiology models such as the Baranyi-Roberts model while providing better explainability.
The inclusion of additional metabolic processes or larger number of cell-species can be achieved by extending the list of possible states and defining transitions between them.
As an example, the description of the quorum sensing phenomena can be modelled by adding the equation for the released autoinducer molecules as a part of the microenvironment.
These molecules then affect the same-species bacteria growth controlling the size of the population.
The approach described above provides a primary predictive model with transition rates between different pools as parameters.
To include the influence of external factors, i.e., temperature, pH, pressure, etc., one can simply use the secondary models for the transition rates, e.g., the square-root temperature dependence~\citep{ratkowsky_relationship_1982}.\\
By constructing an \ac{abm}, we have exemplified how close the devised model is to the individual biological 
processes that occur within the bacteria.
This mechanistic single-cell level of understanding of our \ac{abm} is a perspective which can not be easily replicated by other spatial models such as \acp{pde}.
We have investigated the impact of spatial effects in the form of varying diffusion parameters and different initial spatial distributions of bacteria.
Our findings reinforce that the \ac{ode} model is a good simplification in the highly-diffusive regime within a mostly homogenous spatial bacterial distribution.
However, deviations to these assumptions can lead to non-negligible effects, influencing the growth of the colony.

% ------------------------------------------------------------------------------------------------
\section{Conclusion}
% \begin{itemize}
%     \item New theoretical framework
%     \item consistent with \ac{ode} models (Baranyi-Roberts)
%     \item Better explainability
%     \item Include effects such as inhibition, toxins in understandable manner
%     \item showed spatial limitations of \ac{ode} model
%     \item \ac{abm} model allowed modeling of cellular processes; ensure that pool model is derivable
%         from the individual behvaiour of cellular processes
%     \item outlook: generalized mathematical formulation; extend inhibition model (because it is a
%         dimensional reduction of existing models)
%     \item Provide code in a pypi package as well (TODO)
% \end{itemize}

We introduced the pool model as a framework to unify existing approaches and allow researchers to follow a mechanistic approach during model development.
It is able to capture the effects of many established models and provides a lens which is useful to generalize existing systems and results in better overall explainability.
By modeling effects such as mutual inhibition on a single-cell level within an \ac{abm}, we have exemplified that the pool model directly builds on these more fundamental concepts.
It can thus aid in the mechanistic understanding of the underlying processes, allowing researchers to deepen their knowledge about microbial growth within food science.
Going forward, this framework can provide a starting point for a more abstract mathematical formulation which should be able to cover a much wider class of systems.

@@

% ================================================================================================
\newpage
\beginsupplement
% ================================================================================================
\section{General mathematical formalism}
\label{ssec:supplement1}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, all discussed in the main part systems  can be described with the Ordinary Differential Equations (\acp{ode}) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
\label{eq:model_ODE_x}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector describing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentration of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.\\
%
The advantage of the following formulation is its generality.
Depending on the knowledge about the system as well as the needed level of accuracy one can define an arbitrary number of states $M$ and control the complexity of the model.
Moreover, opposite to the already known Predictive Microbiology like Baranyi and Roberts one can easily change.\\
%
To calculate the bacterial count for one species $i$, we simply need to sum the tensor/vector $\mbx$ over all possible states:
\begin{equation}
    N_i = \sum_k \mbx_i^k.
\label{eq:N_pro_species}
\end{equation}
While the total bacterial count of all species of our system is
\begin{equation}
    N = \sum_i \sum_k \mbx_i^k.
\label{eq:N_total}
\end{equation}
%
%
%The three pool system described with equations~\ref{eq:ode_3pools} will be written as
%\begin{equation}
%    \dot \mbx  = A \mbx, \hspace{3mm} A = \begin{pmatrix}
%        -(\mu + \lambda) & 0             & 0      \\
%        \lambda          & \alpha - \mu' & 0      \\
%        \mu              & \mu'          & -\beta 
%    \end{pmatrix}.
%\label{eq:x_repres_3pools}
%\end{equation}
%
Using such matrix representation of the bacterial system, we can rewrite, for example, the three-pool model with limiting resource described with equations~\ref{eq:ode_3pool_resource} in a more compact form.
The concentrations of the pools are then the components of our variable vector $\mbx (t) = (L(t), G(t), D(t))^T$.
%
\begin{align}
    \dot \mbx  &= \mbF(\mbx, \mbm, \mbp, \mbu) = \begin{pmatrix}
        -(\mu + \lambda) & 0               & 0      \\
        \lambda          & \alpha R - \mu' & 0      \\
        \mu              & \mu'            & -\beta 
    \end{pmatrix} \mbx\\
    \dot \mbm &= \mbH (\mbx, \mbm, \mbp, \mbu) = -\frac{\alpha}{N_t} R G.
\end{align}
Separating linear and non-linear parts in two terms yields
\begin{equation}
    \dot \mbx = \begin{pmatrix}
        -(\mu + \lambda) & 0       & 0 \\
         \lambda         & \alpha  & 0 \\
         \mu &  \mu'& - \beta 
    \end{pmatrix} 
    \mbx + \frac{\alpha}{N_t} x^2 \sum_i^3 x^i \begin{pmatrix} 0 \\ 1 \\ 0  \end{pmatrix}.
\label{eq:x_repres_3pools_resource2}
\end{equation}
Here the first term is the linear part and the second term define the quadratic behavior of the system.
%
\paragraph{Generalized Lotka-Volterra}

>>>>

Bacterial consortia are highly complex systems which consist of multiple distinct cell types.
The underlying processes are interconnected with each other and occur on all scales, ranging from
population-based dynamics of colony growth over molecular processes such as metabolism to gene
expression.
Bacterial communities are ubiquitous and even able to survive under extreme conditions, making them
virtually omnipresent around any living system.
They play an especially important role within the context of food safety and production
being a key contributor to the spoilage of food products as well as fermentation processes.
Furthermore, effects such as biofilm formation can hinder drug treatment by shielding the bacterial
colony from antibiotic substances or immune response of the organism.
The natural bacterial processes are a crucial part of our planet's ecosystem, being responsible for
tasks such as nutrient cycling, decomposition of organic matter and oxygen production.
It is of great interest to understand the dynamics of these various processes.\\
In order to achieve a comprehensive description of the target system, it is important to incorporate
all relevant biological phenomena arising in these bacterial systems.
During their reproductive process, bacteria undergo multiple phases~\citep{buchanan_when_1997}, starting in a lag phase which
is necessary to adapt to new environments and initially impedes the overall growth~\citep{rolfe_lag_2012, monod_growth_1949}.
Additionally, bacteria are able to cooperate, e.g., via complementary metabolic processes or inhibit
each other, e.g., by means of waste production as a metabolic byproduct~\citep{west_social_2007, rainey_evolution_2003, hibbing_bacterial_2010, stubbendieck_bacterial_2016}.
A limited supply of nutrients will also affect the overall reproductive capacity of the colony~\citep{monod_growth_1949}.
Bacteria have thus devised various survival strategies to ensure sufficient access to nutrients.
Effects on a single-cell level such as adhesion~\citep{htuson_bacteriasurface_2013}, migration~\cite{decoene_microscopic_2011} and chemotaxis can lead to complex
phenomena on a population-based level which can aid in the goal of survival.
Cell signaling, e.g., in the form of quorum-sensing can lead to collective emergent phenomena such as
biofilm formation which play an important role for many mature microbial communities~\citep{Jin2020,ng_bacterial_2009}.\\
The first step in the construction of any mathematical model is to determine the required level of
detail with respect to the description scale and consequently which effects to include.
The bacterial growth curves let us describe the dynamics of the overall abundance of various
species.
The field of predictive microbiology has developed a variety of ODE-based models to capture
bacterial growth dynamics~\citep{perez-rodriguez_predictive_2012}.
Various \acp{ode} \textit{primary models} describe the bacterial growth curve under constant
environmental conditions.
Widely used examples include the modified Gompertz model~\citep{Gompertz1825-wi,Gibson1987}, Baranyi-Roberts
model~\citep{baranyi_modeling_1993, baranyi_dynamic_1994} and McKellar two-compartment model~\cite{buchanan_when_1997} which distinguishes between growing and non-growing bacteria.
To take into account changing environmental conditions, so-called \textit{secondary models}
formulate the model parameters as a function of external quantities such as temperature, pH or
humidity.
These can be simple polynomial models, square root-type models~\citep{ratkowsky_relationship_1982} or
more complex relations such as the cardinal parameter model~\citep{zwietering_decision_1992}.
The advantage of the \ac{ode} formulation lies in the predictive power and simplicity of usage.
However, due to their coarse-grained nature, they lack the ability to describe heterogeneity and
other spatial effects.\\
When dealing with spatial effects such as heterogeneity, we are required to use more extensive
modeling formulations such as \acp{pde} or \acp{abm}.
The former represent an intermediate step between the population-based \ac{ode} models and the
individually-based \acp{abm} and can be effective, especially when dealing with large crowds of
cells that can be approximated as a continuum.
On the other hand, \acp{abm} excel at describing the individual behavior of cells which can then
lead to more complex collective phenomena.
Tools such as BSim~\citep{Gorochowski2012,Dang2020}, AgentCell~\cite{emonet_agentcell_2005}, BacSim~\cite{kreft_bacsim_1998} or \texttt{cellular\_raza}~\cite{Pleyer2025} are
capable to describe large amounts of bacteria, performing various intracellular and extracellular
processes.
The latter has been used to adapt \ac{pde}-based models into an \ac{abm} formulation which describe
bacterial branching patterns~\citep{Kawasaki1997,Matsushita1998}.\\
In this work, we introduce a new \ac{ode}-based modeling framework which describes different
bacterial species, their metabolic states and external quantities such as nutrients as individual
pools.
The dynamics of bacterial growth can be understood in terms of transitions between the different
bacterial states, i.e., pools.
Our model also allows us to capture interactions between these bacterial states and the external
microenvironment.
Compared against the aforementioned methods, this approach turns out to be less complex with general
applicability, low number of parameters and greater explainability for the underlying biological
reality.
To illustrate the generality of the model, multiple examples are reported.
We analyze the spatial limitations of an interacting two-species model and show that it can be fully
understood on an individual-based level in terms of single-cell processes.
Furthermore, we assess how spatial heterogeneity~\citep{mckellar_heterogeneous_1997} can impact the shape of the growth curves.

@@

@@

@@

@@

% ================================================================================================
\section{Spatial effects}
As described above, the proposed model operates with the total number of the bacterial counts for
different species and states in the system, not accounting for their spatial distributions.
Assuming that the bacterial system is uniformly distributed (sometimes also called 'well-stirred')
we expect the \ac{ode} description to be valid.
However, in reality spatial effects can play a significant role for population survival forcing some
species to develop competitive mechanisms such as cell adhesion to surfaces in the areas rich with
nutrients~\citep{htuson_bacteriasurface_2013}.
Another mechanism is the motility of bacteria, e.g., the flagellates of \textit{E. coli} or
\textit{B. subtilis} which are capable of chemotactic motion driven by oxygen
concentrations~\citep{decoene_microscopic_2011}.
The bacterial social interactions such as release of toxins or public goods can also
significantly affect the spatial structure of the community~\citep{blanchard_bacterial_2015}.
These effects can only be treated within a spatial framework that is able to describe the observed
effects.\\
Compared to other modeling techniques such as \acp{pde}, \acp{abm}~\citep{nagarajan_agent-based_2022-1} force researchers to think in
terms of cellular processes, thus bridging the gap from a
collective (bulk) model to an individual-based (single-cell) description.
Our goal is to provide an \ac{abm} which can be used to derive the given \ac{ode} model as a
coarse-grained case.
In this way, we can show that the bulk-based \ac{ode} model is directly related to the mechanistic
single-cell behavior of the bacteria.
To this end, we consider the system consisting of two bacterial species that can be described by the two internal states: lag phase and growth.
Both of these species grow using the same resource pool $R$ but with different growth constants
$\alpha_{A/B}$ and different lag phase transition rates $\lambda_{A/B}$.
Moreover, species $A$ is able to release a certain antimicrobial component $I$ that shows an
inhibitory effect on species $B$ (amensalism).
We assume that $L,G$ are measured in terms of the overall volume and $R,I$ are measured as total
amount.
The \acp{ode} system of the bacterial population can be written as:
%
%
\begin{align}
    \begin{split}
        \dot{L}_A &= -\lambda_A R L_A\\
        \dot{G}_A &= \lambda_A R L_A + \alpha_A R G_A\\
        \dot{L}_B &= -\lambda_B R L_B\\
        \dot{G}_B &= \lambda_B R L_B + \frac{\alpha_B}{1 + \mu_I I} R G_B.
    \end{split}
    \label{eq:spatial_limit_F}
\end{align}
%
Moreover, the microenvironment of our system contains the resource and the inhibitor molecules
\begin{align}
    \begin{split}
        \dot{R} &= -\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{(1 + \mu_I I) N_t} R G_B \\
        \dot{I} &= \kappa G_A.
    \end{split}
    \label{eq:spatial_limit_H}
\end{align}
\subsection{Constructing an agent-based pool model}
The following sections will demonstrate how a given \ac{ode} system can be translated into a
custom-built \ac{abm} which has only a few more overall parameters.
This procedure is a blueprint which can be followed to construct \acp{abm} for similar
\ac{ode}-based systems.\\
To construct our \ac{abm}, we use the modeling framework \texttt{cellular\_raza}~\citep{Pleyer2025} that allows us to define our bacterial agents from a clean slate.
With this flexibility, researchers are able to construct minimal models using only as many parameters which are necessary.
This ability, to precisely define and control the exact behavior of our cell-agents, enables the following parts.

@@

The presented pool model was used to describe the dynamics of one- and two-species systems containing the main growth stages and interactions between bacteria in the form of cooperation and competition.
Our approach produces results which exhibit similarities to other well-known predictive microbiology models such as the Baranyi-Roberts model while providing better explainability.
The inclusion of additional metabolic processes or larger number of cell-species can be achieved by extending the list of possible states and defining transitions between them.
As an example, the description of the quorum sensing phenomena can be modelled by adding the equation for the released autoinducer molecules as a part of the microenvironment.
These molecules then affect the same-species bacteria growth controlling the size of the population.
The approach described above provides a primary predictive model with transition rates between different pools as parameters.
To include the influence of external factors, i.e., temperature, pH, pressure, etc., one can simply use the secondary models for the transition rates, e.g., the square-root temperature dependence~\citep{ratkowsky_relationship_1982}.\\
By constructing an \ac{abm}, we have exemplified how close the devised model is to the individual biological 
processes that occur within the bacteria.
This mechanistic single-cell level of understanding of our \ac{abm} is a perspective which can not be easily replicated by other spatial models such as \acp{pde}.
We have investigated the impact of spatial effects in the form of varying diffusion parameters and different initial spatial distributions of bacteria.
Our findings reinforce that the \ac{ode} model is a good simplification in the highly-diffusive regime within a mostly homogenous spatial bacterial distribution.
However, deviations to these assumptions can lead to non-negligible effects, influencing the growth of the colony.
\section{Conclusion}
We introduced the pool model as a framework to unify existing approaches and allow researchers to follow a mechanistic approach during model development.
It is able to capture the effects of many established models and provides a lens which is useful to generalize existing systems and results in better overall explainability.
By modeling effects such as mutual inhibition on a single-cell level within an \ac{abm}, we have exemplified that the pool model directly builds on these more fundamental concepts.
It can thus aid in the mechanistic understanding of the underlying processes, allowing researchers to deepen their knowledge about microbial growth within food science.
Going forward, this framework can provide a starting point for a more abstract mathematical formulation which should be able to cover a much wider class of systems.

@@

% ================================================================================================
\newpage
\beginsupplement
% ================================================================================================
\section{General mathematical formalism}
\label{ssec:supplement1}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, all discussed in the main part systems  can be described with the Ordinary Differential Equations (\acp{ode}) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
\label{eq:model_ODE_x}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector describing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentration of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.\\
%
The advantage of the following formulation is its generality.
Depending on the knowledge about the system as well as the needed level of accuracy one can define an arbitrary number of states $M$ and control the complexity of the model.
Moreover, opposite to the already known Predictive Microbiology like Baranyi and Roberts one can easily change.\\
%
To calculate the bacterial count for one species $i$, we simply need to sum the tensor/vector $\mbx$ over all possible states:
\begin{equation}
    N_i = \sum_k \mbx_i^k.
\label{eq:N_pro_species}
\end{equation}
While the total bacterial count of all species of our system is
\begin{equation}
    N = \sum_i \sum_k \mbx_i^k.
\label{eq:N_total}
\end{equation}
%
%
%
Using such matrix representation of the bacterial system, we can rewrite, for example, the three-pool model with limiting resource described with equations~\ref{eq:ode_3pool_resource} in a more compact form.
The concentrations of the pools are then the components of our variable vector $\mbx (t) = (L(t), G(t), D(t))^T$.
%
\begin{align}
    \dot \mbx  &= \mbF(\mbx, \mbm, \mbp, \mbu) = \begin{pmatrix}
        -(\mu + \lambda) & 0               & 0      \\
        \lambda          & \alpha R - \mu' & 0      \\
        \mu              & \mu'            & -\beta 
    \end{pmatrix} \mbx\\
    \dot \mbm &= \mbH (\mbx, \mbm, \mbp, \mbu) = -\frac{\alpha}{N_t} R G.
\end{align}
Separating linear and non-linear parts in two terms yields
\begin{equation}
    \dot \mbx = \begin{pmatrix}
        -(\mu + \lambda) & 0       & 0 \\
         \lambda         & \alpha  & 0 \\
         \mu &  \mu'& - \beta 
    \end{pmatrix} 
    \mbx + \frac{\alpha}{N_t} x^2 \sum_i^3 x^i \begin{pmatrix} 0 \\ 1 \\ 0  \end{pmatrix}.
\label{eq:x_repres_3pools_resource2}
\end{equation}
Here the first term is the linear part and the second term define the quadratic behavior of the system.
%
\paragraph{Generalized Lotka-Volterra}
